---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Fast and accurate randomized singular value decomposition (RSVD)

## Introduction

There are 3 versions of RSVD implemented in this package, which are ordered by their accuracy.

- **winSVD**: [window based randomized singular value decomposition](https://genome.cshlp.org/content/33/9/1599)
- **dashSVD**: randomized singular value decomposition with dynamic shifted eigenvalues
- **sSVD**: [single pass randomized singular value decomposition with power iterations](https://genome.cshlp.org/content/33/9/1599)

<!-- badges: start -->
<!-- badges: end -->


## Installation

``` r
# For the CRAN version
install.packages("pcaone")
# For the latest developing version
## devtools::install_github("Zilong-Li/PCAoneR")
```

## Example

This is a basic example which shows you how to use pcaone:

```{r example}
library(pcaone)
mat <- matrix(rnorm(100*5000), 5000, 100)
res <- pcaone(mat, k = 10)
str(res)
```

## Accuracy


```{r acc}
library(RSpectra) ## svds
library(rsvd)     ## regular rsvd
library(pcaone)
data(popgen)
A <- popgen - rowMeans(popgen) ## center
k <- 40
system.time(s0 <- RSpectra::svds(A, k = k) )
system.time(s1 <- rsvd::rsvd(A, k = k, q = 4))
system.time(s2 <- pcaone(A, k = k, method = "ssvd"))
system.time(s3 <- pcaone(A, k = k, method = "winsvd"))

par(mar = c(5, 5, 2, 2))
plot(s0$d-s1$d, ylim = c(0, 10), xlab = "PC index", ylab = "Error of singular values", cex = 2, cex.lab = 2)
points(s0$d-s2$d, col = "orange", cex = 2)
points(s0$d-s3$d, col = "red", cex = 2)
legend("top", legend = c("rSVD", "sSVD", "winSVD"), pch = 16,col = c("black", "orange", "red"), horiz = T, cex = 2, bty = "n" )

```

## Benchmarking

Let's see the performance of ```pcaone``` compared to the other rsvd packages. 

``` r
library(microbenchmark)
timing <- microbenchmark(
  'RSpectra' = svds(A,k = k),
  'rSVD' = rsvd(A, k=k, q = 7),
  'pcaone.rsvd' = pcaone(A, k=k, p = 7, method = "rsvd"),
  'pcaone.winsvd' = pcaone(A, k=k, p = 7),
  times=5)
print(timing, unit='s')
```

## References

* [Zilong Li, Jonas Meisner, Anders Albrechtsen (2023). Fast and accurate out-of-core PCA framework for large scale biobank data](https://genome.cshlp.org/content/33/9/1599)

## Todo

- [ ] add ```center``` and ```scale``` method
